Computational methods in landscape ecology

### 2024-07-01 Manuscript.tex
Rename files

\subsection{Neutral landscapes models}\label{neutral-landscapes-models}

Neutral landscapes models simulate landscape patterns without assuming
any underlying abiotic or biotic processes. Thus, they are often used as
null hypothesis, baselines, or scenarios for comparisons with real
landscape patterns, while controlling certain aspects of the landscape
(Li et al. 2004; Wang and Malanson 2008). The first neutral models
simply assigned land cover types randomly to cells in the landscape and
were based on percolation theory (Gardner et al. 1987). Further
developments included hierarchical models that considered different
spatial scales while assigning land cover types (O'Neill et al. 1992).
While these first two classes can only simulate landscape with discrete
land cover classes, fractal neutral models can also simulate continuous
land cover maps by utilizing e.g., Brownian motion (Palmer 1992).
Borrowing from computer graphics, more recent neutral landscape models
make use of spectral synthesis (e.g., Perlin noise; Etherington 2022) or
Binary space partitioning (Etherington et al. 2022) hierarchical neutral
landscapes. In order to simulate landscapes dominated by human
activities approaches based on least-cost network movement exist
(Etherington et al. 2024). Neutral landscapes models can also be used to
simulated landscapes similar to real landscapes (Inkoom et al. 2017), or
based on target values of landscape metrics (Van Strien et al. 2016;
Justeau-Allaire et al. 2022).

\subsection{Landscape Generator}\label{landscape-generator}

Processes-based landscape generators are related to neutral landscapes
models because they can simulate virtual landscapes. However, in
contrast to neutral landscape models, processes-based landscape
generators are able to simulate structurally more realistic landscapes
by including pattern creating processes (Langhammer et al. 2019;
Salecker et al. 2019).

@@

@@

>>>>

\subsection{Landscapes generators}\label{landscapes-generators}

Landscape generators are typically used as null hypotheses, baselines,
or scenarios in landscape ecology analyses because they allow to control
certain aspects of the landscape (Li et al. 2004; Wang and Malanson
2008) and are generally simpler as ecological simulations models (Hess
et al. 2020). Landscape generators can be classified into two major
categories, namely pattern-based and process-based approaches (Pe'er et
al. 2013).

Pattern-based landscape generators simulate landscape patterns without
assuming any underlying abiotic or biotic processes (i.e., neutral
landscape model; Gardner et al. 1987). The earliest neutral models are
based on percolation theory and assign LULC types randomly to cells in
the landscape(Gardner et al. 1987) or include hierarchical models that
consider different spatial scales while assigning cell values (O'Neill
et al. 1992). Landscapes characterized by continuous values can be
simulated by fractal models, such as Brownian motion (Palmer 1992).
Borrowing from computer graphics and being able to simulate hierarchical
landscapes, more recent neutral landscape models make use of spectral
synthesis (e.g., Perlin noise; Etherington 2022) or Binary space
partitioning (Etherington et al. 2022). Neutral landscape models are
also able to simulate landscapes dominated by anthropogenic activities,
e.g., based on least-cost networks (Etherington et al. 2024). In order
to ensure realistic landscapes, several approaches exist that use real
landscapes (Inkoom et al. 2017) or are based on target values
optimization (Van Strien et al. 2016; Justeau-Allaire et al. 2022) to
simulate neutral landscapes.

Contrastingly, processes-based landscape generators explicitly include
hypothesized pattern creating processes (Langhammer et al. 2019;
Salecker et al. 2019).

@@

\bibitem[\citeproctext]{ref-Hess2020}
Hess B, Dreber N, Liu Y, et al (2020) PioLaG: A piosphere landscape
generator for savanna rangeland modelling. Landscape Ecology
35:2061--2082. \url{https://doi.org/10.1007/s10980-020-01066-w}

@@

\bibitem[\citeproctext]{ref-Peer2013}
Pe'er G, Zurita GA, Schober L, et al (2013) Simple process-based
simulators for generating spatial patterns of habitat loss and
fragmentation: A review and introduction to the g-RaFFe model. PLoS ONE
8:e64968. \url{https://doi.org/10.1371/journal.pone.0064968}

### 2024-07-01 Sections/05-simulations.qmd
Rename files

## Neutral landscapes models
Neutral landscapes models simulate landscape patterns without assuming any underlying abiotic or biotic processes.
Thus, they are often used as null hypothesis, baselines, or scenarios for comparisons with real landscape patterns, while controlling certain aspects of the landscape [@Li2004; @Wang2008].
The first neutral models simply assigned land cover types randomly to cells in the landscape and were based on percolation theory [@Gardner1987].
Further developments included hierarchical models that considered different spatial scales while assigning land cover types [@ONeill1992].
While these first two classes can only simulate landscape with discrete land cover classes, fractal neutral models can also simulate continuous land cover maps by utilizing e.g., Brownian motion [@Palmer1992].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a] hierarchical neutral landscapes.
In order to simulate landscapes dominated by human activities approaches based on least-cost network movement exist [@Etherington2024].
Neutral landscapes models can also be used to simulated landscapes similar to real landscapes [@Inkoom2017], or based on target values of landscape metrics [@VanStrien2016; @Justeau-Allaire2022].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl). -->

## Landscape Generator
Processes-based landscape generators are related to neutral landscapes models because they can simulate virtual landscapes.
However, in contrast to neutral landscape models, processes-based landscape generators are able to simulate structurally more realistic landscapes by including pattern creating processes [@Langhammer2019; @Salecker2019].

<!--
Cellular automaton
[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
[GradientLand Software](https://doi.org/10.1016/j.ecoinf.2014.12.001)
[PioLaG](https://doi.org/10.1007/s10980-020-01066-w)
-->

>>>>

## Landscapes generators

Landscape generators are typically used as null hypotheses, baselines, or scenarios in landscape ecology analyses because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler as ecological simulations models [@Hess2020].
Landscape generators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013].

Pattern-based landscape generators simulate landscape patterns without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or include hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
Borrowing from computer graphics and being able to simulate hierarchical landscapes, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities, e.g., based on least-cost networks [@Etherington2024].
In order to ensure realistic landscapes, several approaches exist that use real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022] to simulate neutral landscapes.
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl). -->
<!-- QRULE -->

Contrastingly, processes-based landscape generators explicitly include hypothesized pattern creating processes [@Langhammer2019; @Salecker2019].
<!-- are able to simulate structurally more realistic landscapes by including -->

<!--
Cellular automata
[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
[PioLaG](https://doi.org/10.1007/s10980-020-01066-w)
-->

### 2024-07-01 Sections/07-spatial-planning.qmd
Rename files

---
title: "Computational methods in landscape ecology"
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
bibliography: CLER.bib
format:
  pdf:
    keep-tex: true
    number-sections: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
---

# Introduction

>>>>

---
bibliography: CLER.bib
---

<!-- Person responsible: Jung -->

# Spatial planning
